### 2017-06-17 Manuscripts/cog-psych/cogpsych.tex
nolan's final comb completed

The ability to creatively generate new concepts and ideas is among the most fascinating aspects of human cognition, but we do not have a strong understanding of the cognitive processes and representations underlying concept generation. In this paper, we study the generation of new categories using the computational and behavioral toolkit of traditional artificial category learning. Previous work in this domain has focused on how the statistical structure of known categories generalizes to generated categories, overlooking the role of contrast between the known and generated categories as a factor. We report two experiments demonstrating that contrast between what is known and what is created is of fundamental importance for categorization and the creative process. We propose a novel, exemplar-based approach explaining our results, and compare the model's performance to two key alternatives. Our experiments and simulations demonstrate specifically how category contrast influences creative generation, and how it can interact with other constraints to produce different types of creations. Our work also serves as a concrete example of how well-established and highly controlled experimental frameworks (such as those in the field of category learning) can be used to make progress in an intriguing yet understudied domain.

@@

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that broadly resembles the earlier ``Path of Least Resistance'' account. The core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the path of least resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed \citep[both of which are factors in the Path of Least Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items according to their similarity to known members of the target category. The copy-and-tweak model provided a poor account of their results, as the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge it. However, its application is notable as a first step toward understanding theories developed in the creative cognition literature using well-known formal approaches from the categorization literature.

@@

Although it is evident that people are {\em capable} of creating new concepts and categories, it is not entirely clear how new concepts are systematically made different from what is already known. The hierarchical Bayesian model developed by \cite{jern2013probabilistic} assumes that differences between observed and generated categories are only due to random variation. The model assumes that generated categories are sampled from the same underlying domain distribution as observed categories, and will thus share a common distributional structure. The model does not make predictions about the {\em location} of the category within the domain (the perceptual instantiation of category members). Under a strict interpretation of their model, given knowledge of a single category within the domain, the most probable new category to be generated is located in {\em exactly} the same location and possesses an identical distributional structure. However, this is not simply an issue with their model but with a broader class of standard hierarchical Bayesian models \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these models assume that at some point of the latent generative process the same underlying distribution generates all of the categories and thus, any differences between categories are due to {\em noise} and should not be {\em systematic}. The best a standard hierarchical Bayesian model can do at capturing contrast is to assume that the new category is placed uniformly at random over stimulus space. But, this defeats the purpose of a hierarchy as it is ignored when determining a new category location!\footnote{It is plausible that some hierarchical Bayesian model could be created that generates categories different from each other. However, this model would not be a standard application or extension of most pre-existing hierarchical Bayesian models. The generative process would need to include a component that promotes contrast.}

The copy-and-tweak model tested by \cite{jern2013probabilistic} also claims little about how generated categories should contrast with what is already known. In their simulations, the model was only tested on generation after the learner had been exposed to members of the target category, and so the model's ability to generate a new category from scratch was not evaluated. However, the model's generation is based exclusively on similarity to known members of the target category; when there are no members of the target category, generation is presumably random.

@@

Although existing accounts of creative generation broadly overlook the role of category contrast in determining what is novel versus familiar, it was implicitly assumed that learners in previous experiments were {\em successful} in creating new categories. Thus, effects of category contrast should be observable within the experimental results of these studies. To provide a test of the influence of category contrast in creative generation, we conducted a novel analysis of Experiment 3 from \cite{jern2013probabilistic}\footnote{Although \cite{jern2013probabilistic} reported four experiments, we focus on Experiment 3 because their first two experiments tested generation of items into known categories, and their fourth experiment was identical to Experiment 3 but with a restricted generation space.}.

@@

Because each experimenter-defined category in the \cite{jern2013probabilistic} experiment possessed a distinct hue shared by all members of the category, it is sensible that participants might generate a category with a hue distinct from the experimenter-provided categories. If creative generation were influenced by category contrast in this way, the hues of generated categories should be systematically different from those of the experimenter-defined categories. Unfortunately, stimulus hue was encoded and presented in the Hue-Saturation-Value (HSV) color space, which is device-dependent and not perceptually normed such that perceived color similarity corresponds to proximity in the color space (as opposed to a color space such as CIELAB that is device-independent and equidistant sets of points correspond to pairs of colors that have the same perceptual similarity; \citealp{wyszecki1967}). Further, they did not calibrate their monitor, and so we cannot know the precise colors presented to participants. As \cite{jern2013probabilistic} were interested in relations between the saturation and length of examples in generated into novel categories, this is not an issue for their analyses and results. However, these issues pose a significant challenge to evaluating contrast between the experimenter-defined and participant-generated categories along the hue dimension. It is plausible that two uncalibrated monitors could display the same HSV color and the colors be perceived in different color categories (especially for color boundaries that vary over lightness, such as the yellow-brown boundary).

@@

Re-analyzing the results of \cite{jern2013probabilistic} provides preliminary support that contrast plays a role in category generation. Taken alongside the analyses reported by \cite{jern2013probabilistic}, our analysis suggests that generated categories tend to be distinct from {\em and} distributionally similar to what is already known. However, it is worth noting that our analysis is still limited: The color groups defined above are imprecise, and it is not clear that our color grouping is consistent with the psychological color boundaries perceived by participants. While we did obtain similar results using a variety of alternative groupings, the hue dimension used in the \cite{jern2013probabilistic} study does not lend itself straightforwardly to the the computation of similarities, and thus we cannot be certain of whether our coding accurately approximates the psychological space of the stimuli. This precludes traditional applications of categorization models to their data as it is usually necessary to encode objects in psychological space in order to accurately determine the similarity between objects. By consequence, although these results likely indicate that contrast exerts {\em some} influence, they do not precisely describe the nature of that influence. In the sections below, we propose a quantitative framework specifying the role of category contrast in creative generation.

@@

Although, as an exemplar model, one of PACKER's proposals is people represent categories in terms of a collection of stored exemplars, it is worth emphasizing that the underlying nature of human category representations is not the primary concern of our work. The choice to develop PACKER within an exemplar framework reflects the facts that exemplar models are well-studied, strongly theoretically motivated, and dominate much of the theoretical and empirical work in categorization. The focus of our work with PACKER concerns the dual constraints of within- and between-class similarity; it is not difficult to imagine how such constraints may be instantiated using alternative frameworks (e.g., \citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of categorization models, see \citealp{pothoswills2011}).

@@

PACKER (as well as its name) was in part inspired by earlier work from the categorization literature \citep{hidaka2011packing}. They argued that natural categories ``pack'' the values of features such that different categories fill the space with distance between one another, while maintaining items within the same category close together. Inspired by this idea, PACKER proposes that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories: the most desirable generation candidates are similar to members of the target category and not similar to members of contrast categories. This is achieved by aggregating similarity across known exemplars differently according to class membership. The aggregated similarity $a(y,x)$ between generation candidate $y$ and stored exemplars $x$ is given by:

@@

To begin our investigation, we sought to extend the early evidence obtained from our analysis of the \cite{jern2013probabilistic} data, under a variety of learning conditions and using more standard stimulus materials. We used an artificial stimulus design: A two dimensional domain of squares, varying in color and size (see Figure \ref{fig:e1-conditions}, panel A). These dimensions have been used in numerous classification learning studies \cite[e.g.,][]{conaway2016similar,conaway2016generalization,shepard1961learning,nosofsky1994comparing}. Unlike those used in the \cite{jern2013probabilistic} experiments, distance on the physical dimensions aligns more directly with perceptual similarity, allowing us to evaluate the role of category contrast in creative generation more precisely. To extend the evidence provided by the \cite{jern2013probabilistic} data, we tested the effects of category contrast after learning one category from a set of qualitatively distinct category structures, as shown in Figure \ref{fig:e1-conditions}.

Figure \ref{fig:e1-conditions}b-d show the values of exemplar dimensions belonging to the experimenter-defined categories (`A', or `Alpha') that participants were assigned to learn about prior to generating a new category. Each participant learned one of the category types during training. In the `Cluster' type, category A is a tight cluster of examples in the space. Perceptually instantiated, the members of category A might, for example, be large and dark in color. In the `Row' type, category A has a row pattern across the space, varying along one feature but not the other. Thus, its members might all be dark in color but would vary in size. Finally, in the `XOR' type, the experimenter-defined category consists of two clusters separated in opposite corners of the space, conforming to the exclusive-or logical structure (e.g., members are small and dark or large and light).

@@

As noted in the introduction of this paper, the task of generating members of a new category can be aligned with common tasks studied in the categorization literature: Whereas classification consists of predicting an object's category label on the basis of its features, inference consists of predicting an observed feature, given a set of observed features and a category label. Generation thus consists of predicting {\em all} features of an object, given a novel category label. In keeping with this insight, we designed our generation task as an extension of the traditional artificial classification learning paradigm. The task differs from traditional work in creative cognition primarily through the use of an artificial domain, which enables the application of computational models. The use of an artificial domain also requires the addition of a training phase, during which participants learn about the categories in the domain. As a result, unlike most previous studies \citep[e.g.,][]{ward1994structured}, participants in our studies have no experience with the domain before the start of the experiment, and the experimenter-defined categories are not hierarchically structured (as are plants and animals).

@@

To evaluate the role of contrast, we computed the number of times each stimulus was generated, as a function of its average city-block distance from members of the experimenter-defined ``Alpha'' category. These data, shown in Figure \ref{fig:e1-distanceplots}, reveal a clear pattern: Examples that are more distant from members of the experimenter-defined categories are more likely to be generated into a new category. This supports contrast being a fundamental constraint on creative cognition and statistical regularity alone being insufficient.

@@

Figure \ref{fig:e1-distanceplots} also depicts, for each participant, the average distance of members within the generated category ({\em within-category} distance) against the average distance between members of the generated and experimenter-defined category ({\em between-category} distance). The narrow distribution of between-category distances in the XOR condition reflects the widely distributed nature of the experimenter-defined category, reducing the possible distances to the experimenter-defined category. These data reveal a systematic pattern: The majority of participants generated categories with greater between-category distance than within-category distance. That is, members of the generated category tended to be more similar to one another than to members of the experimenter-defined category. To evaluate this claim quantitatively, we conducted t-tests comparing the amount of within- and between- class distance in each condition. All conditions possessed greater between-category distance: Cluster, $t(63) = 11.43$, $p < 0.001$; Row, $t(60) = 13.16$, $p < 0.001$; and XOR, $t(57) = 3.64$, $p < 0.001$.  These results provide further evidence of an effect of category contrast: Participants prefer to generate categories that are dissimilar to the learned category but maintain some level of internal cohesion.

@@

With respect to ranges along each axis (X, Y), the generated categories from each condition tend to reflect the ranges of the experimenter-defined categories. The categories generated in the Cluster condition were less widely distributed along the X-axis compared to Row, $t(123) = 5.61$, $p < 0.001$, and XOR, $t(120) = 2.68$, $p = 0.008$. Categories generated in the XOR condition were also less widely distributed along the X-axis compared to Row, $t(117) = 2.56$, $p = 0.012$. This latter effect was not expected because the experimenter-defined category for XOR and Row had similar X-ranges. However, the key finding is that categories from the Cluster condition tended to be more tightly distributed along the X-axis.

Likewise, categories generated in the Row condition had less Y-axis range compared to Cluster, $t(123) = 4.57$, $p < 0.001$ and XOR, $t(117) = 9.26$, $p < 0.001$, and categories from the Cluster condition had less Y-axis range compared to XOR, $t(120) = 3.95$, $p < 0.001$. As expected, the correlations in the Cluster and Row conditions were not systematically positive or negative ($ps > 0.1$). However, the generated categories in the XOR condition tended to possess {\em negatively} correlated dimensions, $t(57) = 2.04$, $p = 0.046$. This finding is notable, as it is the opposite of what would be expected, assuming learners are emulating the distributional structure of the experimenter-defined class (which possesses perfectly positively correlated features). We believe that the failure to replicate this finding is because participants in \cite{jern2013probabilistic} could differentiate the generated category on a third dimension (hue) to maintain the statistical regularities on the other two dimensions. Although the correlation in the XOR condition is significantly negative, it is clear from the box-plot in Figure~\ref{fig:e1-statsboxes} that it would be inappropriate to make a strong conclusion (e.g., the median is close to zero). However, we can conclude with confidence that there are situations where people do not emulate the distributional structure of the given category. This indicates that there is more to category generation than the distributional structure of other categories in the domain. Further, as we will discuss in more detail in the model-based analysis section, this is expected by our proposal that contrast is a fundamental principle in category generation.

@@

@@

Thus, the core results of Experiment 1 indicate that generated categories can systematically differ from what we would expect based on prior work. The negative (or null) correlations observed in the XOR condition suggests an interesting interaction between contrasting from a given category and emulating statistical properties. That is, the constraints on creative generation imposed by category contrast may not simply influence the {\em location} of generated categories, but also their distributional structure. In Experiment 2, we test this claim more systematically.

\section{Experiment 2}

To test whether category contrast influences the distributional structure of generated categories, we sought to identify conditions in which differences in the distributional structure of generated categories cannot be explained by the distributional structure of the experimenter-defined category. We created two new category types (depicted in Figure \ref{fig:e2-conditions}) that possess identical distributional structure (both are tight clusters of examples with no correlation between features), and only differ in their Y-axis position: the `Bottom' category lies in the bottom-center of the space, and the `Middle' category lies in the center. The distributional equality of these conditions is key to the design of the experiment: If the distributional structure of previously learned categories were the only influence on the generated categories, we should observe no difference in the categories participants generate between these two conditions. Will participants distribute their generated category differently between conditions due only to the differences in the available stimulus space for generating a new category?

@@

If category contrast influences the distributional structure of the categories people generate, then we should observe different types of categories according to the shape of the space that is unoccupied by members of previously learned categories. The difference in the Y-axis position between the Bottom and Middle conditions produces a considerable change to the shape of the unoccupied space. Participants assigned to learn the Bottom category should be less likely to generate exemplars into the lower regions of the stimulus space (as these areas possess greater similarity to members of the Bottom category), preferring instead to distribute exemplars across the upper region of the space. This constraint is lifted in the Middle condition, as the Middle category exemplars are equidistant to the upper and lower regions of the space. Accordingly, participants should be more likely to utilize both of these areas. Thus, if category contrast influences the distributional structure of generated categories, we should observe more participants in the Middle condition that generate examples above {\em and} below the experimenter-defined category.

@@

However, our ability to detect differences using a standard $t$-test between the conditions is, in this case, diminished due to the non-normality of the data (Shapiro-Wilk normality test $W=0.77, p< 0.001$ for the Middle condition and $W=0.85, p < 0.001$ for the Bottom condition). Figure \ref{fig:e2-yranges} depicts the Y-axis position of the exemplars generated within each participant's category. The categories are sorted by overall range, then by condition assignment. These data reveal that there were nearly as many participants who generated categories spanning the entire Y-axis as those who generated categories spanning almost none of the Y-axis. The non-normality of the Y-axis range distributions thus requires that we use a different approach to addressing the experiment's main question.

@@

Because our main prediction concerns the generation of exemplars within the upper and lower regions of the domain, we compared the conditions in terms of the frequency with which participants generated examples above and below the categories. Specifically, we counted the number of participants in each condition who placed at least one `Beta' exemplar on the top and bottom `rows' of the space (the maximum and minimum possible y-axis value, respectively). The resulting contingencies data are shown in Table \ref{table:e2-subset-table}.

Firstly, it should be noted that nearly every participant utilized the top and/or bottom rows: only $10 / 122$ participants generated their category entirely within the interior region. Fisher's Exact Tests comparing the conditions reveal that more Middle participants generated an exemplar in the bottom row, $p < 0.001$, again demonstrating the role of contrast in guiding where exemplars are generated. The conditions did not differ in use of the top of the space, $p = 0.16$, however, more Middle participants placed exemplars in the top {\em and} bottom rows, $p = 0.038$. The latter effect is of interest here, as it indicates that the shape of the unoccupied space exerts some influence on the distributional structure of generated categories: Participants in the Middle condition were more likely to generate a category spanning the entire Y-axis. Thus, distributional structure of the generated categories can be influenced by category contrast alone because the distributional structure of the given categories did not differ between groups.

@@

Experiments 1 and 2 revealed systematic and strong effects of category contrast on creative generation. In this section, we report the results of simulations with formal models aimed at explaining our observations. Specifically, we present simulations from the PACKER model, as well as a `copy-and-tweak' model (discussed in Section \ref{section:copytweak-packer}), defined as a variant of PACKER with the $\gamma$ parameter constrained to be one. The comparison of these two models serves to highlight the explanatory role of contrast within PACKER's framework: If contrast affords little explanatory advantage, then the two accounts should produce an equally strong account. We also present simulations from an implementation of the hierarchical Bayesian model proposed by \cite{jern2013probabilistic}, described in-depth in Appendix \ref{ap:hsampling-definition}. The comparison between the hierarchical Bayesian model and PACKER is meant to emphasize the necessity of contrast and demonstrate that generation cannot be explained entirely through the emulation of distributional structure. Each model has complementary strengths and weaknesses. Whereas PACKER is insensitive to the distributional structure of learned categories (relying only on within- and between-category similarity), the hierarchical Bayesian model generates categories exclusively on the basis of knowledge of how existing classes are distributed.

@@

However, because many stimuli were infrequently generated (such items near members of the `Alpha' category), we cannot simply compute the empirical average, as infrequently generated stimuli would be likely to show artificially strong differences. Instead, we used a Bayesian analysis to estimate the mean $\mu_x$ on the assumption that the scores $x$ for each stimulus are normally distributed with an unknown mean and unknown standard deviation. The conjugate Normal-Inverse Gamma distribution provides a straightforward method for this estimation:

>>>>

The ability to creatively generate new concepts and ideas is among the most fascinating aspects of human cognition, but we do not have a strong understanding of the cognitive processes and representations underlying concept generation. In this paper, we study the generation of new categories using the computational and behavioral toolkit of traditional artificial category learning. Previous work in this domain has focused on how the statistical structure of known categories generalizes to generated categories, overlooking the degree of contrast between the known and generated categories as a factor. We report two experiments demonstrating that contrast between what is known and what is created is of fundamental importance for categorization and the creative process. We propose a novel, exemplar-based approach explaining our results, and compare the model's performance to two key alternatives. Our experiments and simulations demonstrate specifically how category contrast influences creative generation, and how it can interact with other constraints to produce different types of creations. Our work also serves as a concrete example of how well-established and highly controlled experimental frameworks (such as those in the field of category learning) can be used to make progress in an intriguing yet understudied domain.

@@

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that broadly resembles the earlier ``Path of Least Resistance'' account. The core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the Path of Least Resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed \citep[both of which are factors in the Path of Least Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items according to their similarity to known members of the target category. The copy-and-tweak model provided a poor account of their results, as the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge it. However, its application is notable as a first step toward understanding theories developed in the creative cognition literature using well-known formal approaches from the categorization literature.

@@

Although it is evident that people are {\em capable} of creating new concepts and categories, it is not entirely clear how new concepts are systematically made different from what is already known. The hierarchical Bayesian model developed by \cite{jern2013probabilistic} assumes that differences between observed and generated categories are only due to random variation. The model assumes that generated categories are sampled from the same underlying domain distribution as observed categories, and will thus share a common distributional structure. The model does not make predictions about the {\em location} of the category within the domain (the perceptual instantiation of category members). Under a strict interpretation of their model, given knowledge of a single category within the domain, the most probable new category to be generated is located in {\em exactly} the same location and possesses an identical distributional structure. This is not an issue with their model specifically, but with a broader class of standard hierarchical Bayesian models \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these models assume that at some point of the latent generative process the same underlying distribution generates all of the categories and thus, any differences between categories are due to {\em noise} and should not be {\em systematic}. The best a standard hierarchical Bayesian model can do at capturing contrast is to assume that the new category is placed uniformly at random over stimulus space. But, this defeats the purpose of a hierarchy as it is ignored when determining a new category location!\footnote{It is plausible that some hierarchical Bayesian model could be created that generates categories different from each other. However, this model would not be a standard application or extension of most pre-existing hierarchical Bayesian models. The generative process would need to include a component that promotes contrast.}

The copy-and-tweak model tested by \cite{jern2013probabilistic} also claims little about how generated categories should contrast with what is already known. In their simulations, the model was only tested on generation after the learner had been exposed to members of the target category, and so the model's ability to generate a new category from scratch was not evaluated. However, the model's generation is based exclusively on similarity to known members of the {\em target} category; when there are no members of the target category, generation is presumably random.

@@

Although existing accounts of creative generation broadly overlook the role of category contrast in determining what is novel versus familiar, it was implicitly assumed that learners in previous experiments were {\em successful} in creating new categories. Thus, effects of category contrast should be observable within the experimental results of these studies. To provide a test of the influence of category contrast within existing data, we conducted a novel analysis of Experiment 3 from \cite{jern2013probabilistic}\footnote{Although \cite{jern2013probabilistic} reported four experiments, we focus on Experiment 3 because their first two experiments tested generation of items into known categories, and their fourth experiment was identical to Experiment 3 but with a restricted generation space.}.

@@

Because each experimenter-defined category in the \cite{jern2013probabilistic} experiment possessed a distinct hue shared by all members of the category, it is sensible that participants might generate a category with a hue distinct from the experimenter-provided categories. If creative generation were influenced by category contrast in this way, the hues of generated categories should be systematically different from those of the experimenter-defined categories. Unfortunately, stimulus hue was encoded and presented in the Hue-Saturation-Value (HSV) color space, which is device-dependent and not perceptually normed such that perceived color similarity corresponds to proximity in the color space (as opposed to a color space such as CIELAB that is device-independent and equidistant sets of points correspond to pairs of colors that have the same perceptual similarity; \citealp{wyszecki1967}). Further, they did not calibrate their monitor, and so we cannot know the precise colors presented to participants. As \cite{jern2013probabilistic} were interested in relations between the saturation and length of examples in generated into novel categories, these issues do not undercut their analyses and results. However, these issues pose a significant challenge to evaluating contrast between the experimenter-defined and participant-generated categories along the hue dimension. It is plausible that two uncalibrated monitors could display the same HSV color and the colors be perceived in different color categories (especially for color boundaries that vary over lightness, such as the yellow-brown boundary).

@@

Re-analyzing the results from \cite{jern2013probabilistic} provides preliminary support that contrast plays a role in category generation. Taken alongside the analyses reported by \cite{jern2013probabilistic}, our analysis suggests that generated categories tend to be distinct from {\em and} distributionally similar to what is already known. However, it is worth noting that our analysis is still limited: The color groups defined above are imprecise, and it is not clear that our color grouping is consistent with the psychological color boundaries perceived by participants. While we did obtain similar results using a variety of alternative groupings, the hue dimension used in the \cite{jern2013probabilistic} study does not lend itself straightforwardly to the the computation of similarities, and thus we cannot be certain of whether our coding accurately approximates the psychological space of the stimuli. This precludes traditional applications of categorization models to their data as it is usually necessary to encode objects in psychological space in order to accurately determine the similarity between objects. By consequence, although these results likely indicate that contrast exerts {\em some} influence, they do not precisely describe the nature of that influence. In the sections below, we propose a quantitative framework specifying the role of category contrast in creative generation.

@@

Although, as an exemplar model, one of PACKER's proposals is people represent categories in terms of a collection of stored exemplars, it is worth emphasizing that the underlying nature of human category representations is not the primary concern of our work. The choice to develop PACKER within an exemplar framework reflects the facts that exemplar models have been throughly evaluated, are strongly theoretically motivated, and dominate much of the theoretical and empirical work in categorization. The focus of our work with PACKER concerns the dual constraints of within- and between-class similarity; it is not difficult to imagine how such constraints may be instantiated using alternative frameworks (e.g., \citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of categorization models, see \citealp{pothoswills2011}).

@@

PACKER (as well as its name) was in part inspired by earlier work from the categorization literature. Specifically, \cite{hidaka2011packing} argued that natural categories ``pack'' the values of features such that different categories fill the domain space with distance between one another, while keeping items within the same category close together. Inspired by this idea, PACKER proposes that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories: the most desirable generation candidates are similar to members of the target category and not similar to members of contrast categories. This is achieved by aggregating similarity across known exemplars differently according to class membership. The aggregated similarity $a(y,x)$ between generation candidate $y$ and stored exemplars $x$ is given by:

@@

To begin our investigation, we sought to extend the early evidence obtained from our analysis of the \cite{jern2013probabilistic} data, under a variety of learning conditions and using more standard stimulus materials. We used an artificial stimulus design: A two dimensional domain of squares, varying in color and size (see Figure \ref{fig:e1-conditions}, panel A). These dimensions have been used in numerous classification learning studies \cite[e.g.,][]{conaway2016similar,conaway2016generalization,shepard1961learning,nosofsky1994comparing}. Unlike those used in the \cite{jern2013probabilistic} experiments, distance on these physical dimensions aligns more directly with perceptual similarity, allowing us to evaluate the role of category contrast in creative generation more precisely. To extend the evidence provided by the \cite{jern2013probabilistic} data, we tested the effects of category contrast after learning one category from a set of qualitatively distinct category structures, as shown in Figure \ref{fig:e1-conditions}.

Figures \ref{fig:e1-conditions}b-d show the values of exemplar dimensions belonging to the experimenter-defined categories (`A', or `Alpha') that participants were assigned to learn about prior to generating a new category. Each participant learned one of the category types during training. In the `Cluster' type, category A is a tight cluster of examples in the space. Perceptually instantiated, the members of category A might, for example, be large and dark in color. In the `Row' type, category A has a row pattern across the space, varying along one feature but not the other. Thus, its members might all be dark in color but would vary in size. Finally, in the `XOR' type, the experimenter-defined category consists of two clusters separated in opposite corners of the space, conforming to the exclusive-or logical structure (e.g., members are small and dark or large and light).

@@

As noted in the introduction of this paper, the task of generating members of a new category can be aligned with common tasks studied in the categorization literature: Whereas classification consists of predicting an object's category label on the basis of its features, inference consists of predicting an observed feature, given a set of observed features and a category label. Generation thus consists of predicting {\em all} features of an object, given a novel category label. In keeping with this insight, we designed our generation task as an extension of the traditional artificial classification learning paradigm. The task differs from traditional work in creative cognition primarily through the use of an artificial domain, which enables the application of computational models. The use of an artificial domain also requires the addition of a training phase, during which participants learn about the categories in the domain. As a result, unlike most previous studies \citep[e.g.,][]{ward1994structured}, participants in our studies have no experience with the domain before the start of the experiment, and the experimenter-defined categories are not hierarchically structured (as are many natural categories).

@@

To evaluate the role of contrast, we computed the number of times each stimulus was generated, as a function of its average city-block distance from members of the experimenter-defined ``Alpha'' category. These data, shown in Figure \ref{fig:e1-distanceplots}, reveal a clear pattern: Examples that are more distant from members of the experimenter-defined categories are more likely to be generated into a new category. This supports the notion that contrast is a fundamental constraint on creative cognition and that suggests that statistical regularity alone is insufficient.

Figure \ref{fig:e1-distanceplots} also depicts, for each participant, the average distance of members within the generated category ({\em within-category} distance) against the average distance between members of the generated and experimenter-defined category ({\em between-category} distance). The narrow distribution of between-category distances in the XOR condition reflects the widely distributed nature of the experimenter-defined category, reducing the possible distances to members of the participant-generated category. These data reveal a systematic pattern: The majority of participants generated categories with greater between-category distance than within-category distance. That is, members of the generated category tended to be more similar to one another than to members of the experimenter-defined category. To evaluate this claim quantitatively, we conducted t-tests comparing the amount of within- and between- class distance in each condition. All conditions possessed greater between-category distance: Cluster, $t(63) = 11.43$, $p < 0.001$; Row, $t(60) = 13.16$, $p < 0.001$; and XOR, $t(57) = 3.64$, $p < 0.001$.  These results provide further evidence of an effect of category contrast: Participants prefer to generate categories that are dissimilar to the learned category but maintain some level of internal cohesion.

@@

@@

With respect to ranges along each axis (X, Y), the generated categories from each condition tend to reflect the ranges of the experimenter-defined categories. The categories generated in the Cluster condition were less widely distributed along the X-axis compared to Row, $t(123) = 5.61$, $p < 0.001$, and XOR, $t(120) = 2.68$, $p = 0.008$. Categories generated in the XOR condition were also less widely distributed along the X-axis compared to Row, $t(117) = 2.56$, $p = 0.012$. This latter effect was not expected because the experimenter-defined categories for XOR and Row had similar X-ranges. However, the key finding is that categories from the Cluster condition tended to be more tightly distributed along the X-axis.

@@

Likewise, categories generated in the Row condition had less Y-axis range compared to Cluster, $t(123) = 4.57$, $p < 0.001$ and XOR, $t(117) = 9.26$, $p < 0.001$, and categories from the Cluster condition had less Y-axis range compared to XOR, $t(120) = 3.95$, $p < 0.001$. As expected, the correlations in the Cluster and Row conditions were not systematically positive or negative ($ps > 0.1$). However, the generated categories in the XOR condition tended to possess {\em negatively} correlated dimensions, $t(57) = 2.04$, $p = 0.046$. This finding is notable, as it is the opposite of what would be expected, assuming learners are emulating the distributional structure of the experimenter-defined class (which possesses perfectly positively correlated features).

We believe that the failure to replicate this finding is because participants in \cite{jern2013probabilistic} could differentiate the generated category on a third dimension (hue) to maintain the statistical regularities on the other two dimensions. Although the correlation in the XOR condition is significantly negative, it is clear from the box-plot in Figure~\ref{fig:e1-statsboxes} that it would be inappropriate to make a strong conclusion (e.g., the median is close to zero). However, we can conclude with confidence that there are situations where people do not emulate the distributional structure of the given category. This indicates that there is more to category generation than the emulation of distributional structure of other categories in the domain. Further, as we will discuss in more detail in the model-based analysis section, this is expected by our proposal that contrast is a fundamental principle in category generation.

@@

Thus, the core results of Experiment 1 indicate that generated categories can systematically differ from what we would expect based on prior work. The negative (or null) correlations observed in the XOR condition suggests an interesting interaction between contrast with a given category and emulation of statistical properties. That is, the constraints on creative generation imposed by category contrast may not simply influence the {\em location} of generated categories, but also their distributional structure. In Experiment 2, we test this claim more systematically.

\section{Experiment 2.}

@@

To test whether category contrast influences the distributional structure of generated categories, we sought to identify conditions in which differences in the distributional structure of generated categories cannot be explained by the distributional structure of the experimenter-defined category. We created two new category types (depicted in Figure \ref{fig:e2-conditions}) that possess identical distributional structures (both are tight clusters of examples with no correlation between features), as they only differ in their Y-axis position: the `Bottom' category lies in the bottom-center of the space, and the `Middle' category lies in the center. The distributional equality of these conditions is key to the design of the experiment: If the distributional structure of previously learned categories were the only influence on the generated categories, we should observe no difference in the categories participants generate between these two conditions. Will participants distribute their generated category differently between conditions due to the differences in the available stimulus space for generating a new category?

If category contrast influences the distributional structure of the categories people generate, then we should observe different types of categories according to the shape of the space that is {\em unoccupied} by members of previously learned categories. The difference in the Y-axis position between the Bottom and Middle conditions produces a considerable change to the shape of the unoccupied space. Participants assigned to learn the Bottom category should be less likely to generate exemplars into the lower regions of the stimulus space (as these areas possess greater similarity to members of the Bottom category), preferring instead to distribute exemplars across the upper region of the space. This constraint is lifted in the Middle condition, as the Middle category exemplars are equidistant to the upper and lower regions of the space. Accordingly, participants should be more likely to utilize both of these areas. Thus, if category contrast influences the distributional structure of generated categories, we should observe more participants in the Middle condition that generate examples above {\em and} below the experimenter-defined category.

@@

However, our ability to detect differences in Y-Axis range using a standard $t$-test between the conditions is, in this case, diminished due to the non-normality of the data (Shapiro-Wilk normality test $W=0.77, p< 0.001$ for the Middle condition and $W=0.85, p < 0.001$ for the Bottom condition). Figure \ref{fig:e2-yranges} depicts the Y-axis position of the exemplars generated within each participant's category. The categories are sorted by overall range, then by condition assignment. These data reveal that there were nearly as many participants who generated categories spanning the entire Y-axis as those who generated categories spanning almost none of the Y-axis. The non-normality of the Y-axis range distributions thus requires that we use a different approach to addressing the experiment's main question.

@@

Because our main prediction concerns the generation of exemplars within the upper and lower regions of the domain, we compared the conditions in terms of the frequency with which participants generated examples above and below the categories. Specifically, we counted the number of participants in each condition who placed at least one `Beta' exemplar on the top and bottom `rows' of the space (the maximum and minimum possible Y-axis value, respectively). The resulting contingencies data are shown in Table \ref{table:e2-subset-table}.

Firstly, it should be noted that nearly every participant utilized the top and/or bottom rows: only $10 / 122$ participants generated their category entirely within the interior region. Fisher's Exact Tests comparing the conditions reveal that more Middle participants generated an exemplar in the bottom row, $p < 0.001$, again demonstrating the role of contrast in guiding where exemplars are generated. The conditions did not differ in use of the top of the space, $p = 0.16$, however, more Middle participants placed exemplars in the top {\em and} bottom rows, $p = 0.038$. The latter effect is of interest here, as it indicates that the shape of the unoccupied space exerts some influence on the distributional structure of generated categories: Participants in the Middle condition were more likely to generate a category spanning the entire Y-axis. Thus, distributional structure of the generated categories can be influenced by category contrast alone.

@@

Experiments 1 and 2 revealed systematic and strong effects of category contrast on creative generation. In this section, we report the results of simulations with formal models aimed at explaining our observations. Specifically, we present simulations from the PACKER model, as well as a `copy-and-tweak' model (discussed in Section \ref{section:copytweak-packer}), defined as a variant of PACKER with the $\gamma$ parameter constrained to be one. The comparison of these two models serves to highlight the explanatory role of contrast within PACKER's framework: If contrast affords little explanatory advantage, then the two accounts should produce an equally strong account. We also present simulations from an implementation of the hierarchical Bayesian model proposed by \cite{jern2013probabilistic}, described in-depth in Appendix \ref{ap:hsampling-definition}. The comparison between the hierarchical Bayesian model and PACKER is meant to emphasize the necessity of contrast and demonstrate that generation cannot be explained entirely through the emulation of distributional structure. Each model has complementary strengths and weaknesses: Whereas PACKER is insensitive to the distributional structure of learned categories (relying only on within- and between-category similarity), the hierarchical Bayesian model generates categories exclusively on the basis of knowledge of how existing classes are distributed.

@@

However, because many stimuli were infrequently generated (such items near members of the `Alpha' category), we cannot simply compute the empirical average of the difference scores, as infrequently generated stimuli would be likely to show artificially strong differences. Instead, we used a Bayesian analysis to estimate the mean $\mu_x$ on the assumption that the scores $x$ for each stimulus are normally distributed with an unknown mean and unknown standard deviation. The conjugate Normal-Inverse Gamma distribution provides a straightforward method for this estimation:
